---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and run devtools::build_readme(). -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 3.5,
  dpi = 150,
  out.width = "100%"
)
```

# parametricboot

<!-- badges: start -->
[![R-CMD-check](https://github.com/DiogoRibeiro7/parametricboot/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/DiogoRibeiro7/parametricboot/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/DiogoRibeiro7/parametricboot/graph/badge.svg)](https://app.codecov.io/gh/DiogoRibeiro7/parametricboot)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](LICENSE.md)
<!-- badges: end -->

parametricboot automates the **parametric bootstrap** for fitted regression
models. You fit a model as usual; the package simulates new responses from it,
refits the model to each of them, and turns the replicates into bias
estimates, standard errors, confidence intervals, prediction intervals and
diagnostic plots.

Supported models:

| Model | Fitted with | How replicates are generated |
|---|---|---|
| Linear and generalised linear models | `lm()`, `glm()` | `simulate()`, then the original call is re-evaluated |
| Linear and generalised linear mixed models | `lme4::lmer()`, `lme4::glmer()` | `simulate()`, then `lme4::refit()` |
| Cox proportional hazards models | `survival::coxph()` | model-based resampling (Davison & Hinkley, 1997, Algorithm 7.3) |

## Installation

parametricboot is a work in progress: it is not on CRAN, and the interface may
still change. Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("DiogoRibeiro7/parametricboot")
```

## Example

A classic dose-response experiment (Collett, 1991): batches of 20 tobacco
budworm moths of each sex were exposed to six doses of a pyrethroid, and the
number killed was recorded. That is only 12 binomial observations, so how far
can the large-sample theory behind `summary(fit)` be trusted?

```{r example}
library(parametricboot)

budworm <- data.frame(
  sex = rep(c("M", "F"), each = 6),
  dose = rep(c(1, 2, 4, 8, 16, 32), times = 2),
  dead = c(1, 4, 9, 13, 18, 20, 0, 2, 6, 10, 12, 16),
  n = 20
)
fit <- glm(cbind(dead, n - dead) ~ sex + log2(dose), data = budworm, family = binomial())

res <- pb_simulate(fit, n = 1000, seed = 2025)
res
```

`pb_summary_table()` reports, for each coefficient, the bootstrap estimates of
bias, standard error and mean squared error, the actual coverage of the
nominal 95% Wald interval, and a bootstrap confidence interval:

```{r summary}
pb_summary_table(res)
```

Here the news is good: the slope is overestimated by about 2%, a small
fraction of its standard error, and the Wald intervals cover as advertised.
The bootstrap distributions are close to normal:

```{r estimates, fig.alt = "Histograms of 1000 bootstrap estimates of the intercept, the sex effect and the log-dose slope. Each is roughly symmetric and centred near the original estimate, which is marked by a dashed line."}
pb_plot_estimates(res)
```

The same replicates give intervals for any prediction, here the dose-response
curve for male moths:

```{r predictions, fig.alt = "Predicted probability of death for male moths rising from about 0.1 at dose 1 to 0.95 at dose 32, with a shaded 95 percent bootstrap band that is widest at low doses."}
males <- data.frame(sex = "M", dose = seq(1, 32, length.out = 100))
pb_plot_predictions(res, males, x = "dose", type = "response")
```

## Functions

| Task | Functions |
|---|---|
| Draw replicates | `pb_simulate()`, `pb_parallel()`, `pb_drop_warned()` |
| Summarise | `pb_summary_table()`, `pb_key_stats()`, `pb_confint()`, `pb_tidy()`, plus `print()`, `summary()` and `confint()` methods |
| Visualise | `pb_plot_estimates()`, `pb_plot_diagnostics()`, `pb_plot_predictions()` |
| Predict | `pb_predict_boot()` |
| Compare models | `pb_compare_models()` |
| Nonparametric helper | `pb_resample()` |

The [getting-started article](https://diogoribeiro7.github.io/parametricboot/articles/parametricboot.html)
(`vignette("parametricboot")`) is a tour that also covers mixed models, Cox
models, and what to do when some refits misbehave. The full
[function reference](https://diogoribeiro7.github.io/parametricboot/reference/)
is on the package website.

## Getting help and contributing

Please report bugs and request features on the
[issue tracker](https://github.com/DiogoRibeiro7/parametricboot/issues).
Contributions are welcome; see the
[contributing guide](.github/CONTRIBUTING.md). This project follows a
[code of conduct](CODE_OF_CONDUCT.md).

## Citation

```{r citation, echo = FALSE, comment = ""}
print(citation("parametricboot"), style = "text")
```

## License

MIT &copy; Diogo Ribeiro. See [LICENSE.md](LICENSE.md).
