---
title: "Getting started with parametricboot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with parametricboot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  out.width = "100%"
)
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
has_survival <- requireNamespace("survival", quietly = TRUE)
```

## The idea

Standard errors and confidence intervals for regression models usually rest on
large-sample approximations. The **parametric bootstrap** checks, and if
necessary replaces, those approximations by simulation:

1. fit the model to the data;
2. simulate a new response from the *fitted* model, keeping the covariates
   fixed;
3. refit the model to the simulated response;
4. repeat steps 2 and 3 many times.

Because the data-generating mechanism in step 2 is known exactly (it is the
fitted model), the spread of the refitted estimates around the original one
shows how the estimator behaves in samples like yours: whether it is biased,
how variable it is, and whether the textbook intervals achieve their nominal
coverage.

parametricboot runs this loop for `lm()`, `glm()`, `lme4::lmer()`,
`lme4::glmer()` and `survival::coxph()` fits, and provides tools to summarise
the result.

```{r setup}
library(parametricboot)
```

## A dose-response experiment

Collett (1991) describes an experiment in which batches of 20 tobacco budworm
moths of each sex were exposed to six doses of a pyrethroid. The response is
the number of moths killed.

```{r fit}
budworm <- data.frame(
  sex = rep(c("M", "F"), each = 6),
  dose = rep(c(1, 2, 4, 8, 16, 32), times = 2),
  dead = c(1, 4, 9, 13, 18, 20, 0, 2, 6, 10, 12, 16),
  n = 20
)
fit <- glm(cbind(dead, n - dead) ~ sex + log2(dose), data = budworm, family = binomial())
```

`pb_simulate()` draws the replicates. With `seed`, the result is reproducible
and your session's random number stream is left untouched.

```{r simulate}
res <- pb_simulate(fit, n = 500, seed = 2025)
res
```

The model is refitted by re-evaluating its call on the original data, so the
two-column response and the `log2(dose)` term need no special treatment. The
only requirement is that the model was fitted with a `data` argument.

### Summaries

```{r summary}
pb_summary_table(res)
```

* `bias`, `std_error` and `mse` describe the bootstrap distribution around the
  original estimate.
* `coverage` is the proportion of replicates whose own Wald interval contains
  the original estimate, that is, the estimated actual coverage of the
  nominal 95% Wald interval.
* `lower` and `upper` are bootstrap confidence limits.

Four kinds of interval are available, through `pb_confint()` or the usual
`confint()` generic:

```{r confint}
pb_confint(res, type = "percentile")
pb_confint(res, type = "basic")
pb_confint(res, type = "student")
confint(res, level = 0.9, type = "normal")
```

The studentised (bootstrap-t) interval, `type = "student"`, scales every
replicate by its own standard error. It is the most accurate of the four when
standard errors are estimated reliably; for a normal linear model it
reproduces the exact t interval, which the percentile interval does not.

### Plots

`pb_plot_estimates()` shows the bootstrap distributions:

```{r plot-estimates, fig.alt = "Histograms of the bootstrap estimates of each coefficient, with the original estimate marked by a dashed line."}
pb_plot_estimates(res)
```

`pb_plot_diagnostics()` helps to judge whether `n` is large enough. The running
mean should have settled down by the right-hand edge of the plot.

```{r plot-diagnostics, fig.alt = "Bootstrap estimates of the log-dose slope plotted against replicate number, with their running mean settling just above the dashed line at the original estimate."}
pb_plot_diagnostics(res, parameter = "log2(dose)")
```

All plots are ordinary 'ggplot2' objects, and `pb_tidy()` returns the
replicates in long format if you prefer to build your own.

```{r tidy}
head(pb_tidy(res))
```

### Predictions

`pb_predict_boot()` returns one column of predictions per replicate; extra
arguments go to `predict()`.

```{r predict}
new_moths <- data.frame(sex = c("M", "F"), dose = 8)
preds <- pb_predict_boot(res, new_moths, type = "response")
dim(preds)
t(apply(preds, 1, quantile, probs = c(0.025, 0.5, 0.975)))
```

`pb_plot_predictions()` does this for a whole grid and plots the result:

```{r plot-predictions, fig.alt = "Predicted probability of death for male moths against dose, with a shaded 95 percent bootstrap band."}
males <- data.frame(sex = "M", dose = seq(1, 32, length.out = 100))
pb_plot_predictions(res, males, x = "dose", type = "response")
```

## When refits misbehave

Refits that throw an error are dropped, and refits that raise warnings are
counted rather than printed one by one. Both are reported in a single warning,
and the second kind deserves attention. Consider a logistic regression on the
32 rows of `mtcars`:

```{r separation}
small <- glm(vs ~ mpg, data = mtcars, family = binomial())
res_small <- pb_simulate(small, n = 500, seed = 2025)
```

In a few of the simulated data sets the two groups are perfectly separated by
`mpg`. The maximum likelihood estimate does not exist for such data, and
`glm()` returns whatever huge value the iterations stopped at. A handful of
these replicates dominates every mean and standard deviation:

```{r separation-stats}
pb_key_stats(res_small)[c("term", "estimate", "bias", "std_error")]
```

`pb_drop_warned()` excludes the replicates that raised warnings:

```{r separation-dropped}
clean <- pb_drop_warned(res_small)
clean
pb_key_stats(clean)[c("term", "estimate", "bias", "std_error")]
```

This reveals the genuine, and substantial, small-sample bias of logistic
regression: both coefficients are overestimated in absolute value by about a
fifth. Two caveats apply.

* The cleaned summaries are conditional on the fit being well behaved. Report
  the proportion of dropped replicates with them.
* Percentile intervals are far less sensitive than means: a quantile registers
  that a few replicates were extreme, but not by how much. They do not need
  this treatment, and are arguably more honest without it:

```{r separation-confint}
pb_confint(res_small)
pb_confint(clean)
```

The studentised interval is even less affected, because a separated refit
combines a huge estimate with an even larger standard error, so that its
studentised value is unremarkable:

```{r separation-student}
pb_confint(res_small, type = "student")
pb_confint(clean, type = "student")
```

Replicates are never dropped automatically, because not every warning signals
a useless fit. `res$warned` flags the affected replicates so that you can
inspect `res$replicates` before deciding.

## Mixed models

For 'lme4' models, new random effects are drawn for every replicate, exactly
as `lme4::bootMer()` does, and the variance components are tracked alongside
the fixed effects. Refitting mixed models is slow, so this is where
`pb_parallel()` pays off: it takes the same arguments as
`pb_simulate()`, plus `workers`, and returns an identical result for a given
`seed`.

```{r glmer, eval = has_lme4}
cbpp <- lme4::cbpp
gm <- lme4::glmer(
  cbind(incidence, size - incidence) ~ period + (1 | herd),
  data = cbpp, family = binomial()
)

res_gm <- pb_simulate(gm, n = 100, seed = 1)
# Equivalent, on two cores: pb_parallel(gm, n = 100, workers = 2, seed = 1)

pb_summary_table(res_gm)
```

The last row is the standard deviation of the herd effects, named as 'lme4'
names it in `confint(gm, oldNames = FALSE)`; correlations appear as `cor_...`
and the residual standard deviation of a linear mixed model as `sigma`.
Variance components are where Wald-type inference is weakest, and 'lme4' does
not even report standard errors for them. That is why their `coverage` is
`NA`, and why studentised intervals are not available for these rows. The
bootstrap shows that the herd standard deviation is underestimated (its
`bias` is negative; with 1500 replicates instead of the 100 used here to keep
this vignette fast, it settles at about 10% of the estimate). This is a
well-known property of maximum likelihood with few groups, here 15 herds:

```{r glmer-variance, eval = has_lme4, fig.alt = "Histogram of the bootstrap estimates of the herd standard deviation, centred slightly below the original estimate marked by a dashed line."}
pb_plot_estimates(res_gm, parameter = "sd_(Intercept)|herd")
```

To test whether a variance component is needed at all, see `pb_lrt()` below.
Pass `re.form = NA` to `pb_predict_boot()` for population-level predictions.

## Cox models

A Cox model leaves the baseline hazard and the censoring mechanism
unspecified, so there is no fully parametric model to simulate from. For
`coxph` fits, `pb_simulate()` uses the model-based resampling algorithm of
Davison and Hinkley (1997, Algorithm 7.3):

* failure times are drawn from the fitted survivor function
  $\hat S_0(t)^{\exp(x^\top\hat\beta)}$, with $\hat S_0$ based on the Breslow
  estimate of the baseline hazard;
* subjects who were censored keep their censoring time, and the others get a
  censoring time drawn from the Kaplan-Meier estimate of the censoring
  distribution, conditional on exceeding their observed time.

In a model with `strata()`, each stratum has its own baseline hazard,
censoring distribution and end of follow-up. Only right-censored, unweighted
models are supported, without `tt()`, penalised or `frailty()` terms,
strata-by-covariate interactions or `cluster()`.

```{r coxph, eval = has_survival}
library(survival)
cox <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung)

res_cox <- pb_simulate(cox, n = 200, seed = 1, keep_fits = FALSE)
pb_summary_table(res_cox)
```

Here the bootstrap standard errors are close to the partial-likelihood ones,
shown below, and the Wald intervals have close to nominal coverage, which is
reassuring. `keep_fits = FALSE` stores only the estimates, which saves memory
when you do not need predictions.

```{r coxph-se, eval = has_survival}
sqrt(diag(vcov(cox)))
```

## Testing nested models

`pb_lrt()` is the parametric bootstrap version of the likelihood-ratio test.
It simulates from the fitted *null* model, refits both models to every
simulated response, and compares the observed statistic with the resulting
reference distribution rather than with a chi-squared distribution.

Does the effect of dose differ between the sexes?

```{r lrt}
separate <- update(fit, . ~ sex * log2(dose))
pb_lrt(fit, separate, n = 500, seed = 2025)
```

With only 12 observations the chi-squared p-value is somewhat too small, as
is typical, though the conclusion is the same here. For the normal linear
model, where the exact answer is known, the bootstrap p-value converges to
that of the F test.

The difference becomes dramatic when the null hypothesis lies on the boundary
of the parameter space, as when testing whether a variance component is zero.
The models may be of different classes, so the null can simply be the model
without the random effect:

```{r lrt-boundary, eval = has_lme4}
dye <- lme4::Dyestuff
no_batch <- lm(Yield ~ 1, data = dye)
batch <- lme4::lmer(Yield ~ 1 + (1 | Batch), data = dye)

test <- pb_lrt(no_batch, batch, n = 500, seed = 2025)
test
```

REML fits are refitted by maximum likelihood, which likelihood-ratio tests
require. The chi-squared p-value is more than three times too large.
`pb_plot_lrt()` shows why: under the null hypothesis the estimated batch
variance, and with it the statistic, is exactly zero in about two thirds of
samples, which no chi-squared distribution can reproduce. The Bartlett
correction only rescales the chi-squared distribution, so it cannot help
either, and here makes matters worse.

```{r lrt-plot, eval = has_lme4, fig.alt = "Histogram of the bootstrap likelihood-ratio statistics with a very tall bar at zero, more than twice as high as the chi-squared approximation drawn in red, and the observed statistic of 5.4 marked far in the right tail."}
pb_plot_lrt(test)
```

## Comparing models that are not nested

`pb_compare_models()` bootstraps several models and reports the sampling
variability of a fit criterion under each of them. Here, three link functions
for the dose-response curve:

```{r compare}
links <- list(
  logit = fit,
  probit = update(fit, family = binomial("probit")),
  cloglog = update(fit, family = binomial("cloglog"))
)
pb_compare_models(links, n = 200, seed = 1)
```

The AIC differences between the links are about one unit, against a sampling
standard deviation (`boot_sd`) of about four: these data cannot tell the links
apart, and a difference in AIC that is small relative to `boot_sd` should not
be over-interpreted.

## Practical advice

* Use at least a few hundred replicates for standard errors and 1000 or more
  for confidence limits; check `pb_plot_diagnostics()`.
* The parametric bootstrap is only as good as the model. It quantifies
  sampling variability *assuming the fitted model is right*; it does not
  detect misspecification. `pb_resample()` offers simple case resampling if
  you want a nonparametric point of comparison.
* Models must be fitted with a `data` argument so that they can be refitted.

## References

Collett, D. (1991) *Modelling Binary Data*. Chapman & Hall.

Davison, A. C. and Hinkley, D. V. (1997) *Bootstrap Methods and their
Application*. Cambridge University Press.

Halekoh, U. and Hojsgaard, S. (2014) A Kenward-Roger approximation and
parametric bootstrap methods for tests in linear mixed models: the R package
pbkrtest. *Journal of Statistical Software*, 59(9), 1-32.
